---
title: "MA5108 Assignment 2 - Answers"
author: "Conor O'Donoghue --  20230671"
date: "October 29, 2020"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# One Way ANOVA Case Study

## 1. Load in CSV

```{r load csv}
# using read.csv is like read.table but header=TRUE and sep=',' by default.
rad_df <- read.csv('./two_way_anova.csv')
head(rad_df)
```

## 2. Use attach() for easy referencing
The attach() function lets us access data with just the column names from the given df, instead of needing to provide the df every time.

```{r attach}
attach(rad_df)
# No more 'rad_df$batch', only 'batch' needed now.
head(batch)
```

## 3. Explain the following code

```{r given code}
rad_df[which(batch=='A'), 1]
```

In the bracket notation, the first arg selects rows, and the second selects columns.

What we want (and what this produces) are the fold change values for all the rows from batch A.

This selects all of the rows where batch is 'A' using the which() function, and then selects only the first column (which is the index of the fold change column). Since we're returning one column, the result is a numeric vector rather than a dataframe, which is what we would get if we included all of the columns instead of just one.

## 4. Creating plots
We need to create a plot showing fold_change data in boxplot format for

i) radiation dose 
ii) each of the four batch values

### Radiation dose boxplot
```{r radiation_dose boxplot}
boxplot(rad_df[which(radiation_dose==2),1],
        rad_df[which(radiation_dose==5),1],
        rad_df[which(radiation_dose==10),1],
        col = 2:4, main="Metabolite Fold-Change Boxplots by Radiation Dose",
        ylab="Metabolite Concentration Fold Change", xlab="Radiation Dose",
        names=c(2, 5, 10))
```

### Batch boxplot
```{r batch boxplot}
boxplot(rad_df[which(batch=='A'),1],
        rad_df[which(batch=='B'),1],
        rad_df[which(batch=='C'),1],
        rad_df[which(batch=='D'),1],
        col = 2:5, main="Metabolite Fold-Change Boxplots by Batch",
        ylab="Metabolite Concentration Fold Change", xlab="Batch",
        names=c("A","B","C","D"))
```

## 5. Comment on boxplots
What we expect is that metabolite concentrations should change with radiation doses, and that whatever site each experiment took place at should have minimal impact on the results (they should essentially be replicates).

As far as radiation dose is concerned, we saw about what we expected. While there's significant overlap between doses of 2 and 5 Gy, the distributions are different for all three, and at a glance it appears that the range of metabolite fold change shrinks and the values lower as radiation dose increases.

What's interesting is what we see when we plot the distributions of fold-change data by site location. We expected sites to have similar distributions as they all had the same number of experiments at each radiation dose, but there appears to be a large difference between the fold change distributions depending on what site the experiments were conducted at.

\newpage
## 6. Use 2 One-Way ANOVA analyses to test both radiation_dose and batch

### Radiation Dose
It's important that R recognizes radiation_dose as a factor with 3 levels (2, 5, and 10) instead of a list of numeric values to do the tests correctly.

Since there are only 3 numeric values, we don't actually need to change them into characters, and can just run as.factor() on radiation_dose as-is.
```{r radiation dose}
as.factor(radiation_dose)
anova(lm(fold_change ~ as.factor(radiation_dose)))
```

Since the p-value associated with our F-statistic is less than 0.05, we can conclude that radiation dose has a significant effect on metabolite concentration fold-change.

### Batch
```{r}
# batch is already a factor, so we don't need to call as.factor()
class(batch)
anova(lm(fold_change ~ batch))
```

Since the p-value associated with the F-statistic for batch as a factor is also less than 0.05, we can conclude that it also has a significant effect on metabolite concentration fold-change.

# Two Way ANOVA Case Study

We're performing a two-way ANOVA to test for batch effects -- we want to determine if the apparent differences between the radiation doses and the batches are actually significantly related.

## 1. Explain why a two-way ANOVA is the correct approach to use

One-Way ANOVA can only test whether a single factor (with multiple levels) has a significant effect.

We can run multiple One-Way ANOVAs for each factor we have, but this will not answer the question as to whether the factors have interactions with each other. This approach assumes that each are completely independent.

With a two-way ANOVA, we get to see whether each factor has a significant effect, and also whether there are any interactions between the factors.

In general, if you have more than one factor, you should always use a two-way ANOVA because you really ought to know if the factors depend on each other or not.

## 2. Implement a Two-Way ANOVA Test

```{r}
anova(lm(fold_change ~ as.factor(radiation_dose) * batch))
```

### 1. Determine the significance of each factor
The F-statistics for each of radiation_dose and batch have associated p-values well below 0.05, so we can conclude that each have a significant effect on metabolite concentration fold-change.

### 2. Determine if there is an interaction effect between factors
While somewhat close, the associated p-value for the interaction effect is not below 0.05, so we can't conclude that the radiation_dose and batch factors have an interaction with one another.

### 3. Comment on the boxplot values previously determined

The boxplots generated for both radiation_dose and batch had clearly different distributions between the levels for each factor. When running an ANOVA test, we're assuming under the null hypothesis that the groups are actually part of the same distribution. If this were actually true, the boxplot for each level within each factor should be very similar to each other, with significant overlap. With radiation dose, 10 Gy had a clearly different distribution from the other two levels, and with batch, locations A and C had clearly different distributions from locations B and D.

### 4. Conclusions on the purpose of ANOVA
We didn't want to know whether the metabolite fold-change was different between radiation doses of 2 and 5, we wanted to know if radiation dose **as a factor** has an effect on metabolite fold-change. Not only that, we had multiple factors we wanted to test for a significant effect on metabolite fold-change, and we also wanted to know if each of these factors had interactions with each other. A Two-Way ANOVA test answers all of these questions for us -- a t-test is for comparing only 2 groups, and a one-way ANOVA is for testing multiple groups but with only a single factor. Since we want to know the effects of multiple factors and their interactions, the two-way ANOVA was the most appropriate test to take.

# Survival Analysis of stem-cell growth factors to limit radiotherapy morbidity

First, we're going to need the survival and spline packages to perform a survival analysis.

```{r}
library(survival)
library(splines)
```


Next we need to load in the csv file and attach it.

```{r load in csv}
sc_df <- read.csv('./irradiation.csv')
head(sc_df)
attach(sc_df)
```

## 1. Compute and compare Kaplan-Meier estimators and test null hypothesis

The null hypothesis assumes that there is no difference between each cohort's survival functions. Using the survival package, we can create a Kaplan-Meier estimator of the survival functions for each cohort using the Surv() function, test the differences between the cohorts (based on Treatment) using survdiff() (which will calculate a log-rank test statistic and give us its associated p-value), and we can plot the survival functions by plotting the output of survfit().

The given csv file only has time and cohort values, so I will assume that there is no right-censorship since I would have no way to determine which time values were censored.

```{r test survival functions}
sdiff <- survdiff(Surv(Time) ~ Treatment)
sdiff
```

```{r plot survival functions}
sfit <- survfit(Surv(Time) ~ Treatment)
sfit
plot(sfit, lty=c(1,2), col=c('black', 'blue'),
     main="Survival Functions by Cohort",
     ylab="Proportion of surviving mice",
     xlab="Time (hours)"
)
legend('topright', c('GF', 'Placebo'),
       lty=c(1,2), col=c('black', 'blue'))
```
The most important part of the above analysis is the p-value derived from survdiff -- since the p-value is below 0.05, we can reject the null hypothesis and conclude that the survival functions between the placebo and intestinal stem cell growth factor cohorts are significantly different.

## 2. Find the times for the 50% morbidity for both cohorts

As shown in the survfit object, the 50% morbidity (aka Median) time value is 25.5 hours for the placebo cohort, and 51.5 hours -- nearly double -- for the intestinal stem cell growth factor cohort.

## 3. Interpret the information on figure 2b from the paper, and compare to the results above.

In the paper, figure 2b shows survival functions for the mice undergoing radiation based on two factors -- treatment (Adlacz or AdRspo) and dose (12 Gy, 14 Gy, or 16 Gy). The curves are steep for all the mice treated with Adlacz and the AdRspo-treated mice given 16 Gy of radiation, with the entire population dying at somewhere around 10 days. However for AdRspo-treated mice below 16 Gy, there were survivors until the end of the experiment at 30 days. Around 1/3 of the mice given 14 Gy survived, and more than half of the mice given 12 Gy survived.

Compared to our results, the placebo mice fared less well than either treatment -- at just 10 Gy, the cohort of placebo mice were all dead at just over 2 days total. The intestinal stem cell growth factor mice, however, had a slower curve down to half its population by the time all of the placebo mice had died, and then just about leveled out for over 100 hours, before steeply dropping just before 200. I don't know exactly why some proportion of their mice survived for the entire test duration of 30 days while all of a sudden the remaining GF-treated mice in our experiment died somewhere around the 8th day (I had thought it might be possible that these are all unlisted dropouts, but then they would be at the exact same time rather than spread out around the ~190 hour mark), but it's clear that the intestinal stem cell growth factor treatment had a significant effect on the time until the mice expired.