---
title: "MA5108_20230671_3.Rmd"
author: "Conor"
date: "1/18/2021"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. RNAseq Analysis
The RNAseq analysis is comparing control and NRDE2-depleted breast cancer cells.

I downloaded the gene_counts .csv file and the Series Matrix .txt file containing the experimental metadata, so now we can load the former into a dataframe.

```{r}
df <- read.csv("GSE119827_Cuff_Gene_Counts.csv")
head(df)
```
Columns AJ1 through AJ6 are the samples (with the first three being the three control replicates, and the last three being the three NRDE2-depleted replicates), but we need to convert the dataframe into a smaller matrix that just contains the expression data.

```{r}
df_exp <- df[c('AJ1', 'AJ2', 'AJ3', 'AJ4', 'AJ5', 'AJ6')]
rownames(df_exp) <- make.names(df$Gene_ID, unique=TRUE)
head(df_exp, 10)
dim(df_exp)
```
Now we have a dataframe only containing the gene counts for the six samples, where each row is a unique gene.
Next we need to find out how many (and which) genes are significantly dysregulated between the conditions, once with and once without a multiple testing correction.

## No multiple testing correction
```{r}
treatment <- c(rep(1, 3), rep(2, 3))
sig_genes = c()
for (i in 1:nrow(df_exp)){
  row <- df_exp[i,]
  values <- as.numeric(row)
  lmobj <- lm(values ~ treatment)
  pval <- summary(lmobj)$coefficients[,4][2]
  if (is.nan(pval)){next}
  if (pval <= 0.05){
    sig_genes <- c(sig_genes, i)
  }
}
```
## Using Panther Database
```{r}
sig_gene_names <- rownames(df_exp)[sig_genes]
write(sig_gene_names, file='sig_genes.txt',
      sep='\n')
```

Alright, we've got a list of significant genes in a text file where each line is the name of a significant gene.

I went to Panther and uploaded the list, and obtained a barchart (screenshot of which will be included in submission).

It seems that the majority of significant genes belonged to the 'binding' or 'catalytic activity' biological function gene ontologies.

## 3. Why?
We're asked to read up on NRDE2, and answer the question as to why the researchers conducted this experiment, and why they used this cell line specifically.

On genecards, NRDE2 (Necessary for RNA Interference, Domain Containing) is a gene involved in RNA degradation and export from the nucleus of the cell, so that it can be transported to the exosomes for degradation.

# Question 2
```{r}
library(survival)
df2 <- read.csv('colon_data.csv')
head(df2)
dim(df2)
unique(df2$rx)
```
## 1. Performance of therapy against placebo
```{r}
sdiff <- survdiff(Surv(df2$time) ~ df2$rx)
sdiff
```
p < 0.05, we can conclude that the three treatments are different.

```{r}
sfit <- survfit(Surv(df2$time) ~ df2$rx)
sfit
plot(sfit, col=c('black', 'blue', 'green'),
     main="Survival by treatment", 
     ylab='Proportion of surviving patients',
     mark.time=c(TRUE, TRUE))

legend('topright', c('Obs', 'Lev', 'Lev+5FU'), lty=c(1,1), col=c('black', 'blue', 'green'))
```

## 2. Effect of gender on survival
### a) placebo
```{r}
df2.obs <- df2[which(df2$rx=="Obs"),]
head(df2.obs)
dim(df2.obs)
```

```{r}
sdiff <- survdiff(Surv(df2.obs$time) ~ df2.obs$sex)
sdiff
```
With a p value of 0.2, it appears that sex had no impact on survival in the placebo cohort.

```{r}
sfit <- survfit(Surv(df2.obs$time) ~ df2.obs$sex)
sfit
plot(sfit, col=c('black', 'blue'),
     main="Survival by sex in placebo cohort", 
     ylab='Proportion of surviving patients',
     mark.time=c(TRUE, TRUE))

legend('topright', c('female', 'male'), lty=c(1,1), col=c('black', 'blue'))
```

### b) just Levamisole
```{r}
df2.lev <- df2[which(df2$rx=="Lev"),]
head(df2.lev)
dim(df2.lev)
sdiff <- survdiff(Surv(df2.lev$time) ~ df2.lev$sex)
sdiff
```
With p=0.3, sex doesn't appear to have an effect on survival in the levamisole-treated cohort either.
My guess is that it won't matter in the cohort that received both levamisole and the placebo either.

```{r}
sfit <- survfit(Surv(df2.lev$time) ~ df2.lev$sex)
sfit
plot(sfit, col=c('black', 'blue'),
     main="Survival by sex in Levamisole cohort", 
     ylab='Proportion of surviving patients',
     mark.time=c(TRUE, TRUE))

legend('topright', c('female', 'male'), lty=c(1,1), col=c('black', 'blue'))
```


### c) both drugs
```{r}
df2.lev5fu <- df2[which(df2$rx=="Lev+5FU"),]
head(df2.lev5fu)
dim(df2.lev5fu)
sdiff <- survdiff(Surv(df2.lev5fu$time) ~ df2.lev5fu$sex)
sdiff
```
p=0.2. Again, nowhere near 0.05, so we can't conclude that sex has any effect on survval in the cohort treated with both drugs either.

```{r}
sfit <- survfit(Surv(df2.lev5fu$time) ~ df2.lev5fu$sex)
sfit
plot(sfit, col=c('black', 'blue'),
     main="Survival by sex in both drugs cohort", 
     ylab='Proportion of surviving patients',
     mark.time=c(TRUE, TRUE))

legend('topright', c('female', 'male'), lty=c(1,1), col=c('black', 'blue'))
```


## 3. The impact of having more than 4 positive lymph nodes for both treatment regimens
Essentially will be the same analysis as the previous section, but with the node4 data column as opposed to the sex column.

### a) placebo
```{r}
sdiff <- survdiff(Surv(df2.obs$time) ~ df2.obs$node4)
sdiff
```
The p value is extremely significant, indicating that whether there were 4 or more positive lymph nodes has a significant effect on survival.

```{r}
sfit <- survfit(Surv(df2.obs$time) ~ df2.obs$node4)
sfit
plot(sfit, col=c('black', 'blue'),
     main="Survival by node4 in placebo cohort", 
     ylab='Proportion of surviving patients',
     mark.time=c(TRUE, TRUE))

legend('topright', c('4 or less', 'more than 4'), lty=c(1,1), col=c('black', 'blue'))
```

### b) just Levasimole

```{r}
sdiff <- survdiff(Surv(df2.lev$time) ~ df2.lev$node4)
sdiff
```

Again, a very significant p-value indicates that there is a significant impact on survival in having more than 4 positive lymph nodes.

```{r}
sfit <- survfit(Surv(df2.lev$time) ~ df2.lev$node4)
sfit
plot(sfit, col=c('black', 'blue'),
     main="Survival by node4 in Levamisole cohort", 
     ylab='Proportion of surviving patients',
     mark.time=c(TRUE, TRUE))

legend('topright', c('4 or less', 'more than 4'), lty=c(1,1), col=c('black', 'blue'))
```

### c) both drugs
```{r}
sdiff <- survdiff(Surv(df2.lev5fu$time) ~ df2.lev5fu$node4)
sdiff
```
While not as strongly significant as the previous p-values, the resulting p-value is still well below 0.05, indicating a strong association between survival and whether or not there are more than 4 positive lymph nodes.

```{r}
sfit <- survfit(Surv(df2.lev5fu$time) ~ df2.lev5fu$node4)
sfit
plot(sfit, col=c('black', 'blue'),
     main="Survival by node4 in both drugs cohort", 
     ylab='Proportion of surviving patients',
     mark.time=c(TRUE, TRUE))

legend('topright', c('less than 4', '4 or more'), lty=c(1,1), col=c('black', 'blue'))
```

#### Comments on node4 values
Unlike sex, it appears that whether or not the patient has more than 4 positive sentinel lymph nodes proximal to the resection location for treatment has a significant impact on survival.

As described in the literature presented in the exam, the more positive sentinal lymph nodes, the poorer the prognosis. The number of positive lymph nodes is a measure of the progression of the cancer -- a higher number indicates that the cancer is more metastatic and is in a further stage. When studying the effect on the drugs on survival, it would likely be a good idea to control for the number of positive lymph nodes, as the presence of more or less patients with more than 4 positive lymph nodes in each group may confound the results of the study between drug treatment and survival.